---
title       : An app for pricing US Representatives
subtitle    : don't hire an expensive lobbyist if a competent bag man will do
author      : Gabi Huiber
job         : consultant
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [shiny,mathjax] # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides

--- 

<style>
em {
  font-style: italic
}
</style>

## The motivation

* [Lawmakers Who Upheld NSA Phone Spying Received Double the Defense Industry Cash](http://www.wired.com/2013/07/money-nsa-vote/).

---

## The working example

*  On July 24, 2013, shortly after the first wave of Snowden revelations, US Representative Justin Amash (R, MI) introduced the "Defund the NSA" amendment.
* It failed 217 to 205, in line with the wishes of:
 - the President of the United States (on record  [here](http://www.businessinsider.com/amash-amendment-nsa-white-house-obama-veto-2013-7))
 - security and defense industry interests (the surveillance state is a customer)
 - US citizens who don't think that the NSA is doing anything wrong, and whose representatives carried the day (this is just how the system should work)

--- .class #id 

## Some questions you might ask

1. So what if No votes collected more from defense interests than Yes votes did? Do US representatives really respond to funding from special interests?
1. If they do, is there a way to get an idea of how much might do the job?
3. Could you use the recipe for pricing votes on other issues where interested parties might be curious?

---

## The model

* With data from [here](http://maplight.org/us-congress/bill/113-hr-2397/1742215/contributions-by-vote?sort=asc&order=$%20From%20Interest%20Groups%3Cbr%20/%3EThat%20Opposed&party[D]=D&party[R]=R&party[I]=I&vote[AYE]=AYE&vote[NOE]=NOE&vote[NV]=NV&voted_with[with]=with&voted_with[not-with]=not-with&state=&custom_from=01/01/2011&custom_to=12/31/2012&all_pols=1&uid=44999&interests-support=&interests-oppose=D2000-D3000-D5000-D9000-D4000-D0000-D6000&from=01-01-2011&to=12-31-2012&source=pacs-nonpacs&campaign=congressional) and logistic regression, you can estimate the probability of a No vote as a function of party affiliation and industry funding.
* You can do this under two assumptions: 
 - there's only a baseline difference but representatives from either party respond to funding the same way: 
 $$
 P(No) \sim f(Party + Amount)
 $$
 - there's a difference both at baseline and in how representatives from each party respond to funding: 
 $$
 P(No) \sim f(Party * Amount)
 $$


--- 

## Interpreting the model

* The first assumption mirrors the sentiment of every third-party candidate, whether Green, Socialist, or Libertarian, and of every disengaged citizen, that the two main parties are interchangeable; you may be sympathetic, but if the second assumption fits the data better, that's evidence that you're wrong.
* If under either assumption the response is flat over some plausible range of funding, that is evidence that special interests are wasting their time: when the vote does go their way, it's simply that what they want matches what the majority wants. 
* In other words, if you see no increase in the probability of voting your way as you increase the funding, don't bother trying to find the right price: the politicians are not for sale.

---

## The quantity of interest

* We will run a logit under each assumption above, but won't report the parameter estimates. They're not what's of interest.
* Instead, we will plot the *expected probability of a No vote over a reasonable range of funding levels*, by party, and simulate the uncertainty surrounding it.
* This will immediately tell an interested party how much they should expect to pay for a vote in their favor, with the probabillity of such a vote shown within a chosen confidence band -- 80%, 95%, etc.
* This is easy to do with [Zelig](http://zeligproject.org/); the mechanics were explained [here](http://gking.harvard.edu/files/abs/making-abs.shtml) in 2001(!)

---

## The proof of concept

<iframe src=https://ghuiber.shinyapps.io/Amash/></iframe>

---

## Don't read too much into it

* The Amash amendment was a great candidate for a simple proof-of-concept. 
 - It failed narrowly; this produced balanced classes, which helps the logistic regression classifier fit response curves within narrow probability ranges. 
 - But it did not fail along party lines; this left enough of the variation in `Vote` to be explained by `Amount`, rather than `Party`.
 - The Yes side got no funding. This made things a little easier to reason about.
 - This was an ideologically polarizing issue, but not a strictly partisan one;  law-and-order Republicans and blue dog Democrats went against old-school hipppie Democrats and libertarian-leaning Republicans. 
 - This amendment was probably not very likely to pass. If that's how it was perceived going into the voting, it was more like a photo opp where individual representatives would be happy to cross party lines and score easy points for integrity, on either side.
 - When they get both funding and cover, US Representatives can probably be relied on to be responsive.

--- 

## More interesting use cases
 
* You could find an issue with money on both sides and ideological differences less stark where there's voting and MapLight data available, and run a version of the model that can price both Yes and No outcomes. It's an easy tweak.
* Or you could find an issue with money on both sides but very stark partisan differences, where this app will show you that the battle is not worth fighting: there's no reasonable range of funding where individual representatives will cross party lines. 
* [H.R. 37 - Promoting Job Creation and Reducing Small Business Burdens Act](http://maplight.org/us-congress/bill/114-hr-37/6586030/total-contributions) might be the perfect example of the latter: Republicans loved it, Democrats hated it, and special interests funded generously both sides.

---

## A battle too big for this app

<iframe src=https://ghuiber.shinyapps.io/hr37/></iframe>

---

```{r read_data, cache=TRUE, echo=FALSE}
knitr::read_chunk('get_data.R')
```

## How could you have told, before running the app?
```{r get_data, eval=TRUE, echo=FALSE}
```

* Amash votes crossed party lines:
```{r amash_tab,eval=TRUE,echo=FALSE}
table(amash$Party,amash$Vote)
```

* HR 37 votes, not so much:
```{r hr37_tab,eval=TRUE,echo=FALSE}
table(hr37$Party,hr37$Vote)
```

---

## Thank you

The code is on [GitHub](https://github.com/ghuiber/Amash)
